---
title: "Reproducible Research Project"
output: html_document
---


# loading and prepocessing the data
```{r, echo = TRUE} 
# download file from web
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip", mode="wb")
# unzip data and read 
unzip("activity.zip")
stepdata <- read.csv("activity.csv", header = TRUE)
head(stepdata)
```





# Find mean total number of steps each day.


### Calculate the total number of steps taken each day
```{r, echo = TRUE}
dailysteps <- aggregate(steps ~ date, data = stepdata, sum)
summary(dailysteps$steps)
```


### Make a Histogram of the total number of steps taken each day
```{r, echo = TRUE}

library(ggplot2)

dailystepplot <- ggplot(data =  dailysteps, aes(x= date, y= steps))  +            
    geom_bar(stat="identity", position="dodge") + 
    ggtitle("Total Number of Steps Per Day") +
    theme(axis.text.x=element_blank()) +
    xlab("Day") +   
    ylab("Steps") 
dailystepplot
```


### Calculate and report the mean and median of the total number of steps taken per day
```{r, echo = TRUE}

mean(dailysteps$steps) 
median(dailysteps$steps) 

```





# What are the daily activity patterns?


### Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days 
```{r, echo = TRUE}
steptime <- aggregate(steps ~ interval, data = stepdata, mean)

qplot(interval, steps, data = steptime) + geom_line() +
    ggtitle("Mean Number of Steps Per 5 Minute Interval") 

```


### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}
tail(steptime[order(steptime$steps),])
# interval 835 has greatest amount of steps (206.2)
```





# Imputing missing values
#### Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.


### Calculate and report the total number of missing values in the dataset 
```{r, echo = TRUE}
totaldata <- read.csv("activity.csv", header = TRUE, na.strings = "NA")
valuesonly <-na.omit(totaldata)
nrow(valuesonly) - nrow(totaldata)
# 2304 missing values in the dataset.
```


### Devise a strategy for filling in all of the missing values in the dataset. 

### Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo = TRUE}
meansteps <- aggregate(steps ~ interval, data = stepdata, mean)
cleandata <- totaldata 
for (i in 1:nrow(cleandata)) {
    if (is.na(cleandata$steps[i])) {
        cleandata$steps[i] <- meansteps[which(cleandata$interval[i] == meansteps$interval), ]$steps
    }
}

head(cleandata)
## no longer see NA values in head in correspondin intervals (see first table)
```


### Make a histogram of the total number of steps taken each day.
```{r, echo = TRUE}
cleandailysteps <- aggregate(steps ~ date, data = cleandata, sum)

cleandailyplot <- ggplot(data = cleandailysteps, aes(x= date, y= steps))  +         
    geom_bar(stat="identity", position="dodge") + 
    ggtitle("Total Number of Steps Per Day, Filled in Values.") +
    theme(axis.text.x=element_blank()) +
    xlab("Day") +   
    ylab("Steps") 
cleandailyplot
```


### Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r, echo = TRUE}

mean(cleandailysteps$steps) #Mean steps per = 10766
median(cleandailysteps$steps) #Median steps per day = 10766

```
#### These values do not have any real significant differences from values of first part of assignment.
#### Imputing missing data did not impact the minimum or maximum steps per day, but it did have an impact on it's distribution.
##### The imputed data has quartile values closer to the mean, as you would suspect following the law of large numbers (imputing created a larger dataset)

### Imputed data
```{r, echo = TRUE}
summary(cleandailysteps$steps) 
```


### Original data with missing values
```{r, echo = TRUE}
summary(dailysteps$steps) 
```         





# Are there differences in activity patterns between weekdays and weekends?


### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r, echo = TRUE}
cleandata$date <- as.Date(cleandata$date, format = "%Y-%m-%d")

cleandata$day <- factor(weekdays(cleandata$date))

levels(cleandata$day) <- list(weekday = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
                              weekend = c("Saturday", "Sunday"))
head(cleandata)
```
```{r, echo = TRUE}
table(cleandata$day)
```


### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r, echo = TRUE}
library(lattice)

steptime <- aggregate(steps ~ interval + day, data = cleandata, mean)

xyplot(steptime$steps ~ steptime$interval | steptime$day, 
    layout = c(1, 2), type = "l", main = "Average daily steps, Weekend VS. Weekday (imputed data)",
    xlab = "Interval", ylab = "Number of steps")
 
```